Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-22 RevisedSubmission/hines_hetland_2017_gji.tex
did more

%% DEFINE MATH MODE MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\pos}{\vec{x}} % position vector
\newcommand{\data}{\mitbf{d}} % data vector
\newcommand{\post}{\hat{u}} % posterior
\newcommand{\points}{\mitbf{P}} % observation points
\newcommand{\strain}{\dot\varepsilon} % strain rate
\newcommand{\E}[1]{\e\left[ #1 \right]} % Expected value
\newcommand{\Cov}[1]{\mathrm{Cov}\left[ #1 \right]} % Expected value
\newcommand{\e}{\mathrm{e}} % east subscript
\newcommand{\n}{\mathrm{n}} % north subscript
\newcommand{\G}{\mitbf{G}} % basis functions
\newcommand{\zeros}{\mitbf{0}} % zeros matrix
\newcommand{\eye}{\mitbf{I}} % identity matrix

@@

@@

@@

@@

Here we propose using Gaussian process regression (GPR)
\citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR
is closely related to kriging \citep{Cressie1992} and least squares
collocation \citep{Moritz1978}. The latter has been used by
\citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain
rates from GNSS data. GPR is a Bayesian, non-parametric method for
inferring a continuous signal from scattered data. Since GNSS stations
are irregularly spaced and observation times may differ between
stations, GPR is an ideal tool for synthesizing discrete GNSS data
into a spatially and temporally continuous representation of surface
deformation. GPR is Bayesian in that we describe our prior
understanding of the geophysical signal with a Gaussian process. A
Gaussian process is a continuous stochastic process whose value at any
collection of points can be described as a Gaussian random vector.
Just as a Gaussian random vector is fully determined by a mean vector
and a covariance matrix, a Gaussian process is fully determined by a
mean function and a covariance function. For example, Brownian motion,
$B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has
the mean function $\E{B(t)} = 0$ and the covariance function
$\Cov{B(t),B(t')} = \min(t,t')$, where $t,t' \ge 0$. If there is no
information available to help choose an appropriate prior Gaussian
process, then maximum likelihood methods can be used to objectively
choose one that is most consistent with the data. Once a prior is
selected, we incorporate GNSS observations with the prior to form a
posterior estimate of transient strain. The posterior transient strain
is also a Gaussian process, and we can use its distribution to
confidently discern geophysical signal from noise. We use GPR to infer
strain resulting from SSEs in the Pacific Northwest, demonstrating
that GPR is an effective tool for detecting transient geophysical
processes.

@@

We seek a spatially and temporally dependent estimate of transient
strain rates. We consider transient strain rates to be any deviation
from secular strain rates. We denote transient strain rates as
$\strain(p)$, where $p$ represents the ordered pair $(\pos,t)$, $\pos
= (x_\e,x_\n)$ are spatial coordinates, and $t$ is time. For
simplicity, we assume that the study region is sufficiently small that
$\pos$ can be considered a point in a 2-D map projection that is
aligned with the cardinal directions. We determine $\strain$ by
spatially and temporally differentiating estimates of transient
displacements, which are constrained by GNSS data.

We let $\vec{u}(p) = (u_\e(p),u_\n(p))$ be our prior understanding of
transient displacements. Since transient displacements are not
precisely known, we do not consider $\vec{u}$ to be a deterministic
function. Instead, $\vec{u}$ is a stochastic process containing a
distribution of functions that could potentially describe transient
displacements. We let each component of $\vec{u}$ be a Gaussian
process with zero mean and a generic covariance function
$\Cov{u_i(p),u_i(p')} = C_{u_i}(p,p')$. Using a more concise notation,
we write our prior on each component of $\vec{u}$ as $u_i \sim
\mathcal{GP}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be
positive definite in order to be a valid covariance function. By
definition, $C_{u_i}$ is a positive definite function if the matrix
$[C_{u_i}(p,p')]_{(p,p') \in \points \times \points}$ is positive
definite for any set of points $\points$ \citep[sec.
2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into
spatial and temporal functions as
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}(p,p') = C_{u_i}\left((\pos,t),(\pos',t')\right) 
              = X_i(\pos,\pos')T_i(t,t').
\end{equation}   
As long as the functions $X_i$ and $T_i$ are positive definite,
$C_{u_i}$ is guaranteed to also be positive definite \citep[sec.
4.2.4]{Rasmussen2006}. We also require that the derivatives
$\partial^2 X_i(\pos,\pos')/ \partial x_j \partial x_j'$ and
$\partial^2 T_i(t,t') / \partial t \partial t'$ exist. This ensures
that $\vec{u}$ is spatially and temporally differentiable, allowing us
to compute $\strain$ (See \citet[sec. 2.2]{Adler1981} or \citet[sec.
10A]{Papoulis1991} for a definition of stochastic differentiation and
the conditions for differentiability).

To give an example of $u_i$, we can use a squared exponential function
for $X_i$ and $T_i$,
\begin{equation}\label{eq:ExampleCov}
X_i(\pos,\pos') = \exp\left(\frac{-||\pos - \pos'||_2^2}{2\ell^2}\right), \ \ \ \ 
T_i(t,t') = \phi^2 \exp\left(\frac{-|t - t'|^2}{2\tau^2}\right),
\end{equation}
which satisfies our requirements for positive definiteness and
differentiability. The parameters $\ell$ and $\tau$ control the
length-scale and time-scale, respectively, of realizations (i.e.,
sample functions) of $u_i$. The parameter $\phi$, which we have
arbitrarily chosen to incorporate into $T_i$ rather than $X_i$,
controls the amplitude of realizations of $u_i$. Ideally, we want
realizations of $u_i$ to have a length-scale, time-scale, and
amplitude that resembles what we expect from true signal. A few
realizations for $u_i$ are shown in Figure XXX. While the squared
exponential function is a commonly used covariance function for GPR,
it is not appropriate for every application. The appropriate choice
for $X_i$ and $T_i$ may vary depending on the geophysical signal we
are trying to describe. To keep this section sufficiently general, we
hold off on specifying $X_i$ and $T_i$ until Section
\ref{sec:SignalModel}, where we estimate transient strain from slow
slip events in the Pacific Northwest.

>>>>

%% DEFINE MATH MODE MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\pos}{\vec{x}} % position vector
\newcommand{\data}{\mitbf{d}} % data vector
\newcommand{\post}{\hat{u}} % posterior
\newcommand{\points}{\mitbf{P}} % observation points
\newcommand{\strain}{\dot\varepsilon} % strain rate
\newcommand{\E}[1]{\mathrm{E}\left[ #1 \right]} % Expected value
\newcommand{\Cov}[1]{\mathrm{Cov}\left[ #1 \right]} % Expected value
\newcommand{\e}{\mathrm{e}} % east subscript
\newcommand{\n}{\mathrm{n}} % north subscript
\newcommand{\G}{\mitbf{G}} % basis functions
\newcommand{\zeros}{\mitbf{0}} % zeros matrix
\newcommand{\eye}{\mitbf{I}} % identity matrix

@@

% Why we care about determining strain

@@

% Existing methods for determining secular strain from GPS

@@

% Existing methods for determining transient strain from GPS

@@

% What we are doing in this paper and why it improves upon existing
% methods

Here we propose using Gaussian process regression (GPR)
\citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR
is closely related to kriging \citep{Cressie1992} and least squares
collocation \citep{Moritz1978}. The latter has been used by
\citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain
rates from GNSS data. GPR is a Bayesian, non-parametric method for
inferring a continuous signal from scattered data. Since GNSS stations
are irregularly spaced and observation times may differ between
stations, GPR is an ideal tool for synthesizing discrete GNSS data
into a spatially and temporally continuous representation of transient
strain. GPR is Bayesian in that we use a prior model to control the
spatial and temporal roughness of the inferred transient strain. The
prior is specified as a stochastic process, namely a Gaussian process.
If there is no information available to help choose an appropriate
prior Gaussian process, then maximum likelihood methods can be used to
objectively choose one that is most consistent with the data. We use
GPR to infer transient strain resulting from SSEs in the Pacific
Northwest, demonstrating that GPR is an effective tool for detecting
transient geophysical processes.

@@

% Overview of what we are trying to do in this section and how

We seek a spatially and temporally dependent estimate of transient
strain rates. We consider transient strain rates to be any deviation
from secular strain rates. We denote transient strain rates as
$\strain(p)$, where $p$ represents the ordered pair $(\pos,t)$, $\pos
= (x_\e,x_\n)$ are spatial coordinates, and $t$ is time. For
simplicity, we assume that the study region is sufficiently small that
$\pos$ can be considered a point in a 2-D Cartesian map projection
that is aligned with the cardinal directions. We determine $\strain$
by spatially and temporally differentiating estimates of transient
displacements (i.e., displacements that deviate from the secular
trend), which are constrained by GNSS data.

% Defining u as a prior Gaussian process

We let $\vec{u}(p) = (u_\e(p),u_\n(p))$ be our prior understanding of
transient displacements. Since transient displacements are not
precisely known, we cannot consider $\vec{u}$ to be a deterministic
function. Instead, $\vec{u}$ is considered to be a stochastic process
containing a distribution of functions that could potentially describe
transient displacements. Specifically, we let each component of
$\vec{u}$ be a Gaussian process. A Gaussian process is a stochastic
process whose value at any collection of points can be described as a
Gaussian random vector. Just as a Gaussian random vector is fully
determined by a mean vector and a covariance matrix, a Gaussian
process is fully determined by a mean function and a covariance
function. For example, Brownian motion, $B(t)$, is a well known
Gaussian process in $\mathbb{R}^1$ which has the mean function
$\E{B(t)} = 0$ and the covariance function $\Cov{B(t),B(t')} =
\min(t,t')$, where $t,t' \ge 0$. We let each component of $\vec{u}$
have zero mean and a generic covariance function $\Cov{u_i(p),u_i(p')}
= C_{u_i}(p,p')$. Using a more concise notation, we write our prior on
each component of $\vec{u}$ as $u_i \sim
\mathcal{GP}\left(0,C_{u_i}\right)$.

% Constraints on the Covariance function

The function $C_{u_i}$ must be positive definite in order to be a
valid covariance function. By definition, $C_{u_i}$ is a positive
definite function if the matrix $[C_{u_i}(p,p')]_{(p,p') \in \points
\times \points}$ is positive definite for any set of points $\points$
\citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be
separated into spatial and temporal functions as
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}(p,p') = C_{u_i}\left((\pos,t),(\pos',t')\right) 
              = X_i(\pos,\pos')T_i(t,t').
\end{equation}   
As long as the functions $X_i$ and $T_i$ are positive definite,
$C_{u_i}$ is guaranteed to also be positive definite \citep[sec.
4.2.4]{Rasmussen2006}. We also require that the derivatives
$\partial^2 X_i(\pos,\pos')/ \partial x_j \partial x_j'$ and
$\partial^2 T_i(t,t') / \partial t \partial t'$ exist. This ensures
that $\vec{u}$ is spatially and temporally differentiable, allowing us
to compute $\strain$ (See \citet[sec. 2.2]{Adler1981} or \citet[sec.
10A]{Papoulis1991} for a definition of stochastic differentiation and
the conditions for differentiability).

We provide an example to give our prior on $\vec{u}$ a more tangible
meaning. we can use a squared exponential function for $X_i$ and
$T_i$,
\begin{equation}\label{eq:ExampleCov}
X_i(\pos,\pos') = \exp\left(\frac{-||\pos - \pos'||_2^2}{2\ell^2}\right), \ \ \ \ 
T_i(t,t') = \phi^2 \exp\left(\frac{-|t - t'|^2}{2\tau^2}\right),
\end{equation}
which satisfies our requirements for positive definiteness and
differentiability. The parameters $\ell$ and $\tau$ control the
length-scale and time-scale, respectively, of realizations (i.e.,
sample functions) of $u_i$. The parameter $\phi$, which we have
arbitrarily chosen to incorporate into $T_i$ rather than $X_i$,
controls the amplitude of realizations of $u_i$. Ideally, we want
realizations of $u_i$ to have a length-scale, time-scale, and
amplitude that resembles what we expect from true signal. A few
realizations for $u_i$ are shown in Figure XXX. While the squared
exponential function is a commonly used covariance function for GPR,
it is not appropriate for every application. The appropriate choice
for $X_i$ and $T_i$ may vary depending on the geophysical signal we
are trying to describe. To keep this section sufficiently general, we
hold off on specifying $X_i$ and $T_i$ until Section
\ref{sec:SignalModel}, where we estimate transient strain from slow
slip events in the Pacific Northwest.
